Improvement of Deterministic NVT Molecular Dynamics via Backward Error Analysis

### 2018-06-07 rigid_bodies.tex
results rev I

\setkeys{acs}{
	abbreviations = false,
	articletitle  = false,
	keywords      = false,
	maxauthors    = 10,
	super         = true
}
% Comment below before submitting
\let\titlefont\undefined
\usepackage[fontsize=11pt]{scrextend}
\usepackage[hidelinks,colorlinks,citecolor=blue]{hyperref}
%\flushbottom
% Comment above before submitting

\usepackage{xr}
\externaldocument{rigid_bodies_supp_info}
\usepackage{booktabs}
\usepackage{threeparttable}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[T1]{fontenc}
\usepackage[table]{xcolor}
\definecolor{lightgray}{gray}{0.9}
\usepackage{tikz}
\usetikzlibrary{decorations.text}
\newcommand{\mt}[1]{\boldsymbol{\mathbf{#1}}}   % matrix symbol
\newcommand{\vt}[1]{\boldsymbol{\mathbf{#1}}}   % vector symbol
\newcommand{\tr}[1]{#1^\text{t}}                % transposition
\newcommand{\diff}[2]{\frac{\partial #1}{\partial #2}} % derivative
\newcommand{\sdiff}[2]{\nabla^2_{#2}#1}         % 2nd derivative
\newcommand{\Ham}[1]{{\mathcal H}_\text{#1}}    % Hamiltonian
\newcommand{\Liu}[1]{i\!L_\text{#1}}            % Liouville
\newcommand{\timestep}{h}
\newcommand{\modified}[1]{\widetilde{#1}}

@@

\title{Improvement of Thermostated Molecular Dynamics via Backward Error Analysis}

@@

Molecular dynamics (MD) simulation involves several types of approximations in the quest for mimicking real systems.
Naively, we tend to rank simulation results by their proximity to experimental data and attribute any disagreement to a lack of accuracy of the force field.
However, systematic errors can possibly be hidden in these deviations, being the cause of non-physical behavior.
For instance, it has been shown \cite{Eastwood_2010} that errors in the discretization of the equations of motion appear to disrupt the equipartition of kinetic energy among molecules of distinct sizes.
A similar artifact occurs for the partition of kinetic energy between translational and rotational degrees of freedom of rigid bodies \cite{Davidchack_2010, Silveira_2017}.
This brings into question the NVT-MD integrator developed by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} in which, assuming translational-rotational equipartition, separate chains of Nos\'{e}-Hoover thermostats are coupled to the two classes of degrees of freedom.
In addition, as we will show later on, this method fails to reproduce the specified temperature even for a time-step size $\timestep$ typically employed in MD.
In fact, \citeauthor{Davidchack_2010} \cite{Davidchack_2010} recommends that the use of separate thermostats for translation and rotation should always be avoided.
Nevertheless, issues arise even for freely-moving particles or when translation and rotation are coupled to a single thermostat.
This occurs because Nos\'{e}-Hoover chains (NHC) \cite{Martyna_1992} and other thermostat methods routinely used in MD are based on the relation between temperature and kinetic energy that is a corollary of the equipartition theorem.

For a symplectic integration of Hamiltonian equations of motion, the breakage of equipartition is arguably only apparent \cite{Eastwood_2010}.
The produced trajectory actually follows the dynamics dictated by a \textit{shadow Hamiltonian} \cite{Tuckerman_2010}, which is no longer separable into exclusively momentum- and position-dependent parts and only coincides with the specified Hamiltonian when $\timestep \to 0$.
Therefore, what seems to occur in fact is a failure to identify the quantity that is subject to equipartition in place of the specified kinetic energy.
As \citeauthor{Eastwood_2010} \cite{Eastwood_2010} have shown, Tolman's generalized equipartition theorem \cite{Tolman_1918, Uline_2008} holds at least for small steps, when such a \textit{shadow kinetic energy} keeps displaying a quadratic dependency on the momenta.
The authors then employed backward error analysis to successfully eliminate the artifacts mentioned above.
Here, we prove that their findings can be extended to the symplectic integration of rigid-body motion.

It seems then reasonable to argue that a NHC thermostat should regulate the shadow kinetic energy instead of the specified, ``physical'' one.
However, the theory underlying the concept of shadow Hamiltonian is only applicable to symplectic integrators, whereas NHC thermostats introduce non-Hamiltonian components to the equations of motion.
In this paper, we devise a new procedure that partially overcomes this limitation.
It consists in using a splitting-based numerical integrator that entails a Hamiltonian component, and then carrying out backward error analysis on this NVE part only.
This allows us to evaluate, with minimal computational overhead, a \textit{refined} kinetic energy for the due replacement in the thermostat equations.
It is worth mentioning that all steps of this procedure are theoretically justified.

Our new method is particularly appealing for rigid-body MD, primarily because discretization errors begin to manifest much earlier than any numerical stability caused by the time-step size.

Rigid-body MD is relevant not only because some molecules are usually modeled as rigid bodies, such as in the ubiquitous case of water \cite{Jorgensen_1983}, but also because it can be employed as a coarse-graining strategy for molecules that can be designed as collection of interconnected rigid bodies, such as proteins and nanoparticles \cite{Knorowski_2012, Patra_2013}.
In fact, the approach of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, which is implemented in software packages such as HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995}, for instance, has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, as well as membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008}, and nanoparticles \cite{Patra_2014}.

The case of rigid water exemplifies this behavior and serves for an excellent test case for our proposal. 
In fact, we are able to eliminate the discretization errors in numerically computed averages for time-step sizes up to 6 fs, outperforming the method introduced by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, as well as our implementation for rigid bodies of the numerical scheme developed by \citeauthor{Martyna_1996} \cite{Martyna_1996}.

@@

In the following, we analyze the performance of the proposed NVT integrators introduced in Secs. \ref{sec:mttk}-\ref{sec:refined_method}, and  compare the results with those obtained employing the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}.  It should be noted that all the simulations were carried out considering $t_d = 100$ fs.
The system under study corresponds to 903 TIP3P\cite{Jorgensen_1983} water molecules at a density of 970 kg/m$^3$.
The 12-6 Lennard-Jones and damped Coulombic interactions were truncated at  10 {\AA} and smoothed via a switching function over the range of 9.5 {\AA} to  10 {\AA}.\cite{Silveira_2017} 
The damping of Coulombic interactions was done by a factor erfc($\alpha r$), with $\alpha$ = 0.29 {\AA}$^{-1}$.

Before showing our main results, it may be instructive first to give a picture of the problem in controlling the temperature when discretization errors are non-negligible. 
The most direct way (the most senstive indicator of the presence of discretization errors) to determine the importance of these errors is by examining the departure of the ``modified Hamiltonian'' $\modified{\Ham{}}$ from the physical Hamiltonian $\Ham{}$\cite{Engle_2005}. 
Fortunately, with the analytical form of $\modified{\Ham{}}$ derived in this paper we are able to perform such an analysis for the NVE integrator which is essential, as will be shown below, to understand the results of the NVT methods. 
In Fig.~\ref{fig:nve} we depict the time evolution of both $\modified{\Ham{}}$ (dotted lines) and $\Ham{}$ (solid lines) for the water system undergoing NVE dynamics with different time-step sizes ($\timestep$).
The numerical integrator correspond to Eq.~\ref{eq:splitting NVE} with the exact solution of free rotation. 
As expected, $\modified{\Ham{}}$ is very well conserved and close to $\Ham{}$ for $\timestep$ fs, but fairly rapidly separates from $\Ham{}$ when increasing $\timestep$. 
It should be remarked that $\modified{\Ham{}}$ corresponds to a simple approximation and, most important, it is computationally very cheap. As a result of employing such approximation, we have larger fluctuations in $\modified{\Ham{}}$ for $\timestep$ = 3 fs and $\timestep$ = 4 fs (add in conclusions that we should use high-order approximations).

\begin{figure}
	\includegraphics{Figures/NVE.eps}
	\caption{Total energy $\Ham{}$ (solid lines) and modified Hamiltonian $\modified{\Ham{}}$ (dotted lines) of 903 TIP3P\cite{Jorgensen_1983} water molecules obtained by NVE MD simulations for different time-step sizes $\timestep$. The numerical integrator correspond to Eq.~\ref{eq:splitting NVE} with the exact solution of free rotation.}
	\label{fig:nve}
\end{figure}

For our implementation for rigid bodies of the numerical scheme due to Martyna \textit{et al.} \cite{Martyna_1996} (Eq.~\ref{eq:splitting NVT}), in Fig.~\ref{fig:props_k_m} we depict the computed averages of temperature ($\langle T \rangle$) and potential energy per molecule ($\langle U \rangle$) as a function of the time-step size. 
The specified temperature corresponds to T = 298 K.
Fig.~\ref{fig:props_k_m} clearly shows that a suitable temperature control leads to a severely degraded accuracy in $\langle U \rangle$, which systematically increases with $\timestep$.
As demonstrated in Fig.~\ref{fig:nve}, the discretization errors are non-negligible even for a small step size of 2 fs.
Thus, the results of Fig.~\ref{fig:props_k_m} can be interpreted as follows: if the kinetic energy is restricted to adopt a certain value, in accordance with the specified temperature, the unavoidable discretization errors will accumulate in the remaining degrees of freedom.
These observations motivated the development of the scheme of Sec.~\ref{sec:refined_method}, whose promising results we show below.

\begin{figure}
	\includegraphics{Figures/thermodynamic_properties_intro.eps}
    \caption{Effect of the time-step size on the thermodynamic properties and of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing...}
	\label{fig:props_k_m}
\end{figure}

@@

The explicit inclusion of the ``refined'' Hamiltonian in a computationally cheap NVT integrator is achieved at the cost of losing the time-reversibility.
This poses the question whether the symmetry loss will influence the long-term stability.
In this paper, we define the energy drift, denoted by $R$, as the the rate of change of energy with time, which corresponds to the angular coefficient obtained from linear regression fitting.
Recall that the quantity expected to be conserved along the simulation is $\widetilde{H}$ (Eq.~\ref{eq:this extended energy}) for our method, which from now on will be referred as ``This work''.
On the other hand, the symmetric scheme due to Martyna \textit{et al.} \cite{Martyna_1996} should preserve $H$ (Eq.~\ref{eq:nvt extended energy}).
In Fig.~\ref{fig:energy_drift}, we present the computed $R$ as a function of the time-step size, and one can see that for these numerical schemes the energy starts to drift upwards from $h$ = 5 fs. 
Also, the long-term stability is only modestly improved by our method so, evidently, the lack of symmetry has a negligible impact on $R$.
Fig.~\ref{fig:energy_drift} also contains the drift in $\Ham{}$ computed for the NVE method\cite{Silveira_2017}, which results to be almost identical to our method.
It is worth mentioning that the stability threshold for the NVE method corresponds to $h$ = 6 fs.
The NVT methods discussed so far allow to extend that limit to 10 fs (see Table~\ref{table:stability} in the Supplementary Material).
We have computed $R$ also for $\Ham{}$, which is the contribution to the extended energy from the physical degrees of freedom, namely the NVE part of the integrator.
In this case, $\Ham{}$ remains bounded for all the time-step sizes considered, so the thermostat is stabilizing the numerical integration, which is also true for the scheme of Martyna \textit{et al.} \cite{Martyna_1996} (see Table~\ref{table:stability} of Supp. Material).
This has already been noticed by Davidchack\cite{Davidchack_2009}, and previously by Bond \textit{et al.} \cite{Bond_2007}, who in this respect stated that ``the thermostat acts as a sort of reservoir for numerical error''.

\begin{figure}
	\includegraphics{Figures/energy_drift.eps}
    \caption{Effect of the time-step size of 903 TIP3P \cite{Jorgensen_1983} water molecules in NVT MD simulations. $R$ for $\widetilde{H}$ and $\Ham{}$, which is the contribution to the extended energy from the physical degrees of freedom. We included the results for the integrator due to Martyna \textit{et al.} \cite{Martyna_1996}
}
	\label{fig:energy_drift}
\end{figure}

To end this section, we present in Fig.~\ref{fig:num_stab} the anomalous behavior encountered when using the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
Particularly, one can see from Fig.~\ref{fig:num_stab} (a) that for each time-step size the extended energy drifts upwards until reaching a point at which the integrator becomes stable.
It should be remarked that these results were double-checked by using the LAMMPS software package \cite{Plimpton_1995}. 
As already mentioned, the scheme due to Kamberaj \textit{et al.} \cite{Kamberaj_2005} couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom.
Considering a factorization scheme different to the one employed by those authors, Davidchack \cite{Davidchack_2009} also implemented such a separate thermostatization and discovered the existence of energy flows from one thermostat to the other.
To the best of our knowledge, the work of Davidchack \cite{Davidchack_2009} did not result in a revision in the literature of the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
As can be seen in Fig.~\ref{fig:num_stab} (b), where we depict the energy of each thermostat for different time-step sizes, that flux is also observed for the scheme due to Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
In addition, the negative values of energy correspond to the thermostat coupled to the translational degrees of freedom, so the energy is flowing from this thermostat to the one coupled to the rotational degrees of freedom.
Interestingly, Fig.~\ref{fig:num_stab} (b) suggests that a unique steady state is established, and the time to reach the steady state decreases with $h$. As far as we know, this behavior has not been reported before.

\begin{figure}
	\includegraphics{Figures/numerical_stability.eps}
    \caption{Total extended energy of 903 TIP3P\cite{Jorgensen_1983} water molecules obtained by NVT MD simulations with different time
step sizes, employing (a) the proposed method represented by Eq. ; (b) our implementation for rigid bodies of the integrator due to Martyna \textit{et al.} \cite{Martyna_1996} and (c) the NVT method introduced by Kamberaj \textit{et al.} \cite{Kamberaj_2005}.Energy of thermostats for 903 TIP3P\cite{Jorgensen_1983} water molecules obtained by NVT MD simulations with different time
step sizes, employing the NVT method introduced by Kamberaj \textit{et al.} \cite{Kamberaj_2005}.}
	\label{fig:num_stab}
\end{figure}

@@

@@

%It is not the purpose of this work to propose an optimal factorization scheme for the NVT dynamics, but rather to prove that a unique thermostat chain is capable of maintaining the average temperature at its specified value while providing the proper sampling of the canonical ensemble.
%First we present some results as first mented.
%That is, the proposed methods coupling to a single thermostat, while in Kamberaj the translational and rotational degrees of freedom are separatedly coupled.
%In Table~\ref{table:new_results}, it can be seen that the integrators proposed here, which are represented by the Eqs.~\eqref{eq:trotter_splitting_NHC} and ~\eqref{eq:modified_splitting}, systemcatically achive the specifed temperature, regardless of the time step size.
%By contrast, the numerical solver introduced by Kamberaj and co-workers\cite{Kamberaj_2005} fails to yield the specified temperature for even the smaller time step size of 1 fs.
%Resultados sobre o checkensemble:

%checkensemble:We report the results corresponding to that method, and since it is instructive to have a graphical insight, we also show some figures of the linear fitting.
%Given that the potential and kinetic energies are separable, the analysis can be made separately as follows.This last approach does not involve any binning of the energies, hence no discretization error is present.
%m-b distribution valid when equipartition of energy is satisfied.

%Recalling that in the NVT dynamics the conserved quantity is the extended energy $H$ of Eq.~\eqref{eq:H_nvt}, in Table~\ref{table:new_results_2} we report the energy drift, denoted by $R$, as the the rate of change of $H$ (kcal/mol) with time (ns), which is the angular coefficient obtained from linear regression fitting  between those variables.
%Except for the case of $\Delta$ = 1 fs, there is also a clear correspondece between an incresing $R$ with the failure in keeping the temperature at its specified value.

%In Fig.~\ref{fig:checkensemble} we show the result from the linear regression analysis for the potential energies sampled using Eq.~\eqref{eq:trotter_splitting_NHC}.
%In figure legends we show the averages temperatures simulated.
%It can be seen that the mentioned algorithm generates a distribution consistent with the canonical ensemble, and that is also true for the kinetic energies.
%Likewise, in Fig.~\ref{fig:mbdistribution} we observe a good correspondence between the kinetic energy distribution at $T= 304.99 K$ and the corresponding Maxwell-Boltzmann distribution We arrive at the same conclusions for the approach given by Eq.~\eqref{eq:modified_splitting}.

%Finally, in Table~\ref{table:ensemblevalidation} we present the results of the maximum likelihood approach applied for both NVT integration schemes.
%It can be seen that the deviations from the true slope, $\timestep$ = 8.970~K, are less than $1\sigma$.
%This means that effectively, the sampled values do not deviate from the correct distribution to a statistically noticeable level.

@@

\bibliography{rigid_bodies}
\end{document}
%A geometric integrator is a numerical method which preserves certain geometric structures of the exact flow of a differential equation.
%the issues arising in the design and analysis of numerical methods for molecular dynamics
%(symplecticness) of the phase flow. We will later show (in Chap. 3) that such symplectic methods preserve a perturbed energy-like invariant (to very high accuracy); the preservation of this modified energy then ensures that the original energy is nearly conserved
%A benefit of respecting the symplectic structure lies in the prospect of using backward error analysis to analyze the behavior of the method
%It seems intuitively correct to demand that a numerical method should mimic as much as possible the available structure of the dynamical system, when this is possible to do without incurring large computational overheads in the process. So much the better that discrete dynamical systems designed in this way appear to be related to modified continuous dynamics with appropriate properties. This observation is purely aesthetic.
%These methods obtain their energy preservation properties indirectly: we design the methods to exactly conserve a certain geometric property
%The purpose of this section is to demonstrate that, by using the principles of geometric integration and the backward error analysis, it is possible to compute effective estimates of the error in numerically computed averages, and even to correct those averages by a straightforward reweighting

>>>>

\setkeys{acs}{
	abbreviations = false,
	articletitle  = false,
	keywords      = false,
	maxauthors    = 10,
	super         = true
}
\usepackage{xr}
\externaldocument{rigid_bodies_supp_info}
% Comment below before submitting
\let\titlefont\undefined
\usepackage[fontsize=11pt]{scrextend}
\usepackage[hidelinks,colorlinks,citecolor=blue]{hyperref}
%\flushbottom
% Comment above before submitting
\usepackage{booktabs}
\usepackage{threeparttable}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[T1]{fontenc}
\usepackage[table]{xcolor}
\definecolor{lightgray}{gray}{0.9}
\usepackage{tikz}
\usetikzlibrary{decorations.text}
\newcommand{\mt}[1]{\boldsymbol{\mathbf{#1}}}   % matrix symbol
\newcommand{\vt}[1]{\boldsymbol{\mathbf{#1}}}   % vector symbol
\newcommand{\tr}[1]{#1^\text{t}}                % transposition
\newcommand{\diff}[2]{\frac{\partial #1}{\partial #2}} % derivative
\newcommand{\sdiff}[2]{\nabla^2_{#2}#1}         % 2nd derivative
\newcommand{\Ham}[1]{{\mathcal H}_\text{#1}}    % Hamiltonian
\newcommand{\Liu}[1]{i\!L_\text{#1}}            % Liouville
\newcommand{\timestep}{h}
\newcommand{\modified}[1]{\widetilde{#1}}

@@

\title{Improvement of Deterministic NVT Molecular Dynamics via Backward Error Analysis}

@@

Molecular Simulation, being the computational realization of Statistical Mechanics \cite{Tuckerman_2010}, involves several types of approximations in order to mimic ``real'' physical systems.
Naively, one would expect those approximations to only produce quantitative deviation from experimental values, and usually we tend to attribute any disagreement to a lack of accuracy of the force field.
However, what can be hidden in the deviations are actually systematic errors, which may introduce non-physical behavior.
For instance, it has already been shown that the unavoidable errors in the discretization of the equations of motion disrupt the equipartition of kinetic energy \cite{Eastwood_2010}.
The same artifact occurs on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body Molecular Dynamics (MD) \cite{Davidchack_2009,Silveira_2017}.
This finding brings into question the NVT integrator developed by Kamberaj and co-workers\cite{Kamberaj_2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom. In addition, as will be shown later, this method fails to reproduce the specified temperature for time-step sizes usually employed in MD.

The relevance of rigid-body MD stems from the fact that not only some molecules are usually modelled as rigid bodies, like the ubiquitous case of water \cite{Jorgensen_1983}, but also it can be employed as a coarse-graining strategy for systems that might be designed as collection of interconected rigid bodies, such as proteins and nanoparticules \cite{Knorowski_2012, Patra_2013}.
As a matter of fact, the approach of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, which is implemented, for instance, in the software packages HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995}, has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, as well as membranes\cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles\cite{Yan_2008} and nanoparticles \cite{Patra_2014}.

The Nos\'{e}-Hoover chain (NHC) thermostat \cite{Martyna_1992}, routinely used in MD, is based on the extended phase-space approach, in which auxiliary degrees of freedom act regulating the kinetic energy of the physical variables so as to be consistent with the Maxwell-Boltzmann distribution at a given target temperature.
This requires an expression relating the temperature, a macroscopic parameter, with the kinetic energy, which is a property of each microstate.
Usually, in MD the equipartition theorem plays this role. 
However, as already mentioned, this theorem is disrupted in numerical schemes with non-negligible discretization errors.
This is because numerical trajectories actually sample from the so-called \textit{shadow} Hamiltonian, which cannot be merely separated into momentum and position dependent parts.
Instead, what is usually identified as the \textit{shadow} kinetic energy contains quadratic terms in particle momenta accompanied by particle positions dependency \cite{Eastwood_2010}.
It seems then reasonable to argue that the thermostat should control this kinetic energy, instead of the physical one.
However, the theory underlying the concept of shadow Hamiltonian correspond to backward error analysis, which is only applicable to symplectic integrators.
This is not the case of NHC numerical schemes, given that the auxiliary variables introduce non-Hamiltonian components to the equations of motion.
In this paper we introduce a procedure that partially overcomes this limitation by performing the backward error analysis on the NVE part of the numerical integrator with minimal computational overheads. 
The resulting NHC scheme, which simultaneously couples the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, controls the fluctuations of an approximation to the \textit{shadow} kinetic energy.

Our new method is particularly appealing for rigid-body MD, primarily because discretization errors begin to manifest much earlier than any numerical stability caused by the time-step size. 
The case of rigid water exemplifies this behavior and serves for an excellent test case for our proposal. 
In fact, we are able to eliminate the discretization errors in numerically computed averages for time-step sizes up to 6 fs, outperforming the method introduced by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, as well as our implementation for rigid bodies of the numerical scheme developed by \citeauthor{Martyna_1996} \cite{Martyna_1996}.

@@

In the following, we analyze the performance of the NVT scheme proposed in  this paper and compare the results with those obtained employing the measure-preserving version of the integrator due to \citeauthor{Martyna_1996} \cite{Martyna_1996} as well as the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}

The system under study corresponds to 903 TIP3P\cite{Jorgensen_1983} water molecules at a density of 970 kg/m$^3$.
The 12-6 Lennard-Jones and damped Coulombic interactions were truncated at  10 {\AA} and smoothed via a switching function over the range of 9.5 {\AA} to  10 {\AA} as done in Ref.~\citenum{Silveira_2017}. 
The damping of Coulombic interactions was done by a factor erfc($\alpha r$), with $\alpha$ = 0.29 {\AA}$^{-1}$.
In addition, all the simulations were carried out considering $t_d = 100$ fs.

Before presenting our main results, we first examine the departure of the \textit{refined} Hamiltonian from the specified one, as this is the most direct way to determine the importance of the discretization erros \cite{Engle_2005}. This is essential for the discussion of the NVT case.

In Fig.~\ref{fig:nve} we show the time evolution of both $\modified{\Ham{}}$ (dotted lines) and $\Ham{}$ (solid lines) for the water system undergoing NVE dynamics with different time-step sizes ($\timestep$). For this, we used the numerical scheme of Eq.~\ref{eq:splitting NVE} with the unsplit solution of rotation. As expected, $\modified{\Ham{}}$ is very well conserved and close to $\Ham{}$ for $\timestep$ = 1 fs, but fairly rapidly separates from $\Ham{}$ when increasing $\timestep$. 
It should be remarked that $\modified{\Ham{}}$ corresponds to a simple approximation and, most important, it is computationally very cheap. As a result of such approximation, we have larger fluctuations in $\modified{\Ham{}}$ for $\timestep$ = 3 fs and $\timestep$ = 4 fs (add in conclusions that we should use high-order approximations).

\begin{figure}
	\includegraphics{Figures/NVE.eps}
	\caption{Total energy $\Ham{}$ (solid lines) and \textit{refined} Hamiltonian $\modified{\Ham{}}$ (dotted lines) for different time-step sizes ($\timestep$) obtained from NVE MD simulations of 903 TIP3P\cite{Jorgensen_1983} water molecules using the numerical scheme of Eq.~\ref{eq:splitting NVE} with the unsplit rotation method.}
	\label{fig:nve}
\end{figure}

@@

The explicit inclusion of the \textit{refined} Hamiltonian in a computationally cheap NVT integrator is achieved at the cost of losing the time-reversibility.
This poses the question whether the symmetry loss will influence the long-term stability.

Here we define the energy drift, denoted by $R$, as the the rate of change of energy with time, which corresponds to the angular coefficient obtained from linear regression fitting.
Recall that the quantity expected to be conserved along the simulation is $\widetilde{H}$ (Eq.~\ref{eq:this extended energy}) for our method, which from now on will be referred as ``This work''.
On the other hand, the symmetric scheme due to Martyna \textit{et al.} \cite{Martyna_1996} should preserve $H$ (Eq.~\ref{eq:nvt extended energy}).
In Fig.~\ref{fig:energy_drift}, we present the computed $R$ as a function of the time-step size, and one can see that for these numerical schemes the energy starts to drift upwards from $h$ = 5 fs. 
Also, the long-term stability is only modestly improved by our method. Evidently, the impact of the lack of symmetry on $R$, if any, is negligible.
Fig.~\ref{fig:energy_drift} also contains the drift in $\modified{\Ham{}}$ for the NVE method \cite{Silveira_2017} (circles), which almost coincides with the results of our method. It is worth mentioning that the stability threshold for the NVE method corresponds to $h$ = 6 fs.
The NVT methods discussed so far allow to extend that limit to 10 fs (see Table~\ref{table:stability} in the Supplementary Material). As one can notice in Fig.~\ref{fig:energy_drift}, the drift in the \textit{refined} Hamiltonian part for our NVT method (squares) remains bounded for all the time-step sizes considered, so the thermostat is stabilizing the numerical integration.
This has already been noticed by Davidchack\cite{Davidchack_2009}, and previously by Bond \textit{et al.} \cite{Bond_2007}, who in this respect stated that ``the thermostat acts as a sort of reservoir for numerical error''.

\begin{figure}
	\includegraphics{Figures/energy_drift.eps}
    \caption{Influence of the time-step size ($h$) on the long-time drift ($R$) in: the extended energy ($\modified{H}$, down triangles) and its \textit{refined} Hamiltonian ($\modified{\Ham{}}$, squares) part for our NVT method, as well as in the extended energy ($H$, up triangles) for the method due to \citeauthor{Martyna_1996} \cite{Martyna_1996} and in the \textit{refined} Hamiltonian ($\modified{\Ham{}}$, circles) for the numerical scheme of Eq.~\ref{eq:splitting NVE} with unsplit solution of rotation.}
	\label{fig:energy_drift}
\end{figure}

To end this section, we present in Fig.~\ref{fig:num_stab} the anomalous behavior encountered when using the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
Particularly, one can see from Fig.~\ref{fig:num_stab} (a) that for each time-step size the extended energy drifts upwards until reaching a point at which the integrator becomes stable.
It should be remarked that these results were double-checked by using the LAMMPS software package \cite{Plimpton_1995}. 
As already mentioned, the scheme due to Kamberaj \textit{et al.} \cite{Kamberaj_2005} couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom.
Considering a factorization scheme different to the one employed by those authors, Davidchack \cite{Davidchack_2009} discovered the existence of energy flows from one thermostat to the other.
To the best of our knowledge, the work of Davidchack \cite{Davidchack_2009} did not result in a revision in the literature of the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
As can be seen in Fig.~\ref{fig:num_stab} (b), where we depict the energy of each thermostat for different time-step sizes, that flux is also observed for the scheme due to Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
In addition, the negative values of energy correspond to the thermostat coupled to the translational degrees of freedom, so the energy is flowing from this thermostat to the one coupled to the rotational degrees of freedom.
Interestingly, Fig.~\ref{fig:num_stab} (b) suggests that a unique steady state is established, and the time to reach the steady state decreases with $h$. As far as we know, this behavior has not been reported before.

\begin{figure}
	\includegraphics{Figures/numerical_stability.eps}
    \caption{Long-time drift in (a) the extended energy and (b) in each thermostat energy for the NVT method introduced by Kamberaj \textit{et al.} \cite{Kamberaj_2005}. In part (b), the negative values correspond to the thermostat coupled to the translational degrees of freedom, so the energy is flowing to the thermostat coupled to the rotational degrees of freedom.}
	\label{fig:num_stab}
\end{figure}

@@

The specified temperature corresponds to T = 298 K.
Fig. clearly shows that a temperature control as described in Sec.~\ref{sec:nvt} leads to a severely degraded accuracy in $\langle U \rangle$, which systematically increases with $\timestep$.
As demonstrated in Fig.~\ref{fig:nve}, the discretization errors are non-negligible even for a small step size of 2 fs.
Thus, the results of Fig.~\ref{fig:props_k_m} can be interpreted as follows: if the kinetic energy is restricted to adopt a certain value, in accordance with the specified temperature, the unavoidable discretization errors will accumulate in the remaining degrees of freedom.
These observations motivated the development of the scheme of Sec.~\ref{sec:refined_method}, whose promising results we show below.

@@

%Given that the potential and kinetic energies are separable, the analysis can be made separately as follows.This last approach does not involve any binning of the energies, hence no discretization error is present.
%m-b distribution valid when equipartition of energy is satisfied.

%It can be seen that the mentioned algorithm generates a distribution consistent with the canonical ensemble, and that is also true for the kinetic energies.
%Finally, in Table~\ref{table:ensemblevalidation} we present the results of the maximum likelihood approach applied for both NVT integration schemes.
%It can be seen that the deviations from the true slope, $\timestep$ = 8.970~K, are less than $1\sigma$.
%This means that effectively, the sampled values do not deviate from the correct distribution to a statistically noticeable level.

@@

\bibliography{rigid_bodies}
\end{document}
